Large-scale feature selection with Gaussian mixture models for the classification of high dimensional remote sensing images

### 2017-01-12 response.tex

\begin{revbox}
  A reference or an explanation should be made in Page 1, line 36 regarding the curse of dimensionality
  \begin{resbox}
    We added the following reference: \emph{D. Donoho, ``High-dimensional data analysis: the curses and blessing of dimensionality,'' in AMS Mathematical challenges of the 21st century, 2000}.
  \end{resbox}
\end{revbox}

\section{Reviewer 3}

\begin{revbox}
  Although the method seems promising, there are some significant issues the authors should address to. The key novelty and contribution of this technique is unfortunately unclear. The authors dedicate a lot of space in explaining already known mathematical formulations of the Gaussian Mixture Models and the feature selection techniques, in comparison with their proposed scheme.
  \begin{resbox}
    We totally agree that some mathematical formulations are conventional. We believe that it is necessary to provide to readers sufficient information to explain clearly the contribution of the proposed paper. The contributions are based on specific derivation of the GMM and the FE formulation. Without these primary recall, we are afraid that readers can be lost. Concerning the contribution of the paper, they are stated in the antepenultimate paragraph of the introduction. They concern the expression of efficient updates rules using block matrix processing. Furthermore, an open source C++ library is made available for the community. The results of our experiment shows that the proposed algorithm has the advantage to be very efficient in term of processing time with a fairly reasonable cost in term of classification accuracy. It has to be noticed that compared to similar methods using Gaussian Mixture model, no additional hypothesis are made on the covariance matrix which is tractable only because of the efficient implementation.

    To make the contribution more clear, the antepenultimate paragraph has been modified to
\rev{The contribution of this paper is a extension of the forward feature selection method proposed in~[27]. A smart implementation of the feature selection update rules are presented in order to perform efficiently on large amount of data. The rules use on linear algebra on block matrices applied to the covariance matrix of the conditional class density function.  Furthermore, a floating version of the algorithm is proposed and evaluated. Several \emph{correctness of fit} criteria are proposed to handle unbalanced training sets, extending the conventional \emph{overall accuracy} measure.  Finally, the developed algorithm is made available to the scientific community as a remote module of the Orfeo Toolbox~[13].
  \end{resbox}
\end{revbox}

### 2017-01-12 response.tex

\begin{revbox}
  A reference or an explanation should be made in Page 1, line 36 regarding the curse of dimensionality
  \begin{resbox}
    We added the following reference: \emph{D. Donoho, ``High-dimensional data analysis: the curses and blessing of dimensionality,'' in AMS Mathematical challenges of the 21st century, 2000}.
  \end{resbox}
\end{revbox}

\section{Reviewer 3}

\begin{revbox}
  Although the method seems promising, there are some significant issues the authors should address to. The key novelty and contribution of this technique is unfortunately unclear. The authors dedicate a lot of space in explaining already known mathematical formulations of the Gaussian Mixture Models and the feature selection techniques, in comparison with their proposed scheme.
  \begin{resbox}
    We totally agree that some mathematical formulations are conventional. We believe that it is necessary to provide to readers sufficient information to explain clearly the contribution of the proposed paper. The contributions are based on specific derivation of the GMM and the FE formulation. Without these primary recall, we are afraid that readers can be lost. Concerning the contribution of the paper, they are stated in the antepenultimate paragraph of the introduction. They concern the expression of efficient updates rules using block matrix processing. Furthermore, an open source C++ library is made available for the community. The results of our experiment shows that the proposed algorithm has the advantage to be very efficient in term of processing time with a fairly reasonable cost in term of classification accuracy. It has to be noticed that compared to similar methods using Gaussian Mixture model, no additional hypothesis are made on the covariance matrix which is tractable only because of the efficient implementation.

    To make the contribution more clear, the antepenultimate paragraph has been modified to
\emph{\rev{The contribution of this paper is a extension of the forward feature selection method proposed in~[27]. A smart implementation of the feature selection update rules are presented in order to perform efficiently on large amount of data. The rules use on linear algebra on block matrices applied to the covariance matrix of the conditional class density function.  Furthermore, a floating version of the algorithm is proposed and evaluated. Several \emph{correctness of fit} criteria are proposed to handle unbalanced training sets, extending the conventional \emph{overall accuracy} measure.  Finally, the developed algorithm is made available to the scientific community as a remote module of the Orfeo Toolbox~[13].}}
  \end{resbox}
\end{revbox}

### 2017-01-13 response.tex

    \emph{For the sake of clarity, only the three aforementioned configurations are discussed here \rev{and the results for all other configurations is available in the supplementary material}.}
  \end{resbox}
\end{revbox}

    \begin{table}[!t]
        \centering
        \caption{Average classification accuracy 20 trials with all configurations of criterion and method (standard deviation in parenthesis).\label{tab:aisa-otbsimu-all}}
        \begin{tabular}{lccc}\toprule
             & \multicolumn{3}{c}{\bfseries Cohen's kappa} \\ \cmidrule{2-4}
            \# samples by class & 250 & 500 & 1000 \\ \midrule

            GMM SFS JM &      {\bfseries 0.685 (0.030)} & {\bfseries 0.689 (0.030)} & {\bfseries 0.693 (0.032)} \\
            GMM SFS divKL &   {\bfseries 0.680 (0.027)} & {\bfseries 0.699 (0.028)} & {\bfseries 0.708 (0.031)} \\
            GMM SFS kappa &   {\bfseries 0.682 (0.025)} & {\bfseries 0.698 (0.027)} & {\bfseries 0.710 (0.029)} \\
            GMM SFS OA &      {\bfseries 0.682 (0.025)} & {\bfseries 0.698 (0.027)} & {\bfseries 0.710 (0.028)} \\
            GMM SFS F1mean &  {\bfseries 0.682 (0.028)} & {\bfseries 0.699 (0.027)} & {\bfseries 0.710 (0.028)} \\
            GMM SFFS JM &     {\bfseries 0.685 (0.030)} & {\bfseries 0.689 (0.030)} & {\bfseries 0.693 (0.032)} \\
            GMM SFFS divKL &  0.547 (0.158) & 0.615 (0.113) & 0.636 (0.120) \\
            GMM SFFS kappa &  0.519 (0.126) & 0.537 (0.102) & 0.545 (0.141) \\
            GMM SFFS OA &     0.531 (0.126) & 0.537 (0.093) & 0.560 (0.127) \\
            GMM SFFS F1mean & 0.447 (0.158) & 0.579 (0.097) & 0.529 (0.129) \\
            \bottomrule
        \end{tabular}
    \end{table}

    \begin{table}[!t]
        \centering
        \caption{Average classification accuracy for 1,000 samples by class averaged over 5 trials  with all configurations of criterion and method (standard deviation in parenthesis).\label{tab:potsdam-otbsimu-all}}
        \begin{tabular}{lccc}\toprule
            & {\bfseries 5\_11 (train)} & {\bfseries 5\_12 (test)} & {\bfseries 3\_10 (test)} \\ \cmidrule{2-4}
            GMM SFS JM &      0.624 (0.028) & 0.631 (0.034) & 0.461 (0.027) \\
            GMM SFS divKL &   0.552 (0.092) & 0.557 (0.086) & 0.388 (0.084) \\
            GMM SFS kappa &   {\bfseries 0.694 (0.002)} & {\bfseries 0.669 (0.005)} & {\bfseries 0.533 (0.008)} \\
            GMM SFS OA &      {\bfseries 0.696 (0.003)} & {\bfseries 0.670 (0.005)} & {\bfseries 0.535 (0.011)} \\
            GMM SFS F1mean &  {\bfseries 0.689 (0.002)} & {\bfseries 0.668 (0.006)} & {\bfseries 0.529 (0.005)} \\
            GMM SFFS JM &     0.624 (0.028) & 0.631 (0.034) & 0.461 (0.027) \\
            GMM SFFS divKL &  0.161 (0.066) & 0.142 (0.065) & 0.148 (0.058) \\
            GMM SFFS kappa &  0.474 (0.103) & 0.499 (0.144) & 0.330 (0.099) \\
            GMM SFFS OA &     0.407 (0.114) & 0.424 (0.148) & 0.335 (0.099) \\
            GMM SFFS F1mean & 0.319 (0.052) & 0.316 (0.067) & 0.316 (0.034) \\
            \bottomrule
        \end{tabular}
    \end{table}

@@

    \begin{table}[!t]
        \centering
        \caption{Average classification accuracy for 1,000 samples by class averaged over 5 trials  evaluated with overall accuracy and mean of F1-score (standard deviation in parenthesis).\label{tab:potsdam-otbsimu-othereval}}
        \begin{tabular}{lcccccc}\toprule
             & \multicolumn{2}{c}{\bfseries 5\_11 (train)} & \multicolumn{2}{c}{\bfseries 5\_12 (test)} & \multicolumn{2}{c}{\bfseries 3\_10 (test)} \\ \cmidrule{2-7}
             & {\bfseries OA} & {\bfseries F1-mean} & {\bfseries OA} & {\bfseries F1-mean} & {\bfseries OA} & {\bfseries F1-mean} \\ \cmidrule{2-7}

            GMM SFS JM &      0.725 (0.025) & 0.528 (0.022) & 0.750 (0.029) & 0.514 (0.023) & 0.585 (0.026) & 0.467 (0.020) \\
            GMM SFS divKL &   0.677 (0.065) & 0.503 (0.044) & 0.702 (0.063) & 0.488 (0.036) & 0.521 (0.071) & 0.414 (0.072) \\
            GMM SFS kappa &   {\bfseries 0.771 (0.002)} & {\bfseries 0.642 (0.005)} & {\bfseries 0.769 (0.004)} & {\bfseries 0.602 (0.004)} & {\bfseries 0.637 (0.007)} & {\bfseries 0.562 (0.010)} \\
            GMM SFS OA &      {\bfseries 0.773 (0.002)} & {\bfseries 0.642 (0.006)} & {\bfseries 0.771 (0.004)} & {\bfseries 0.602 (0.004)} & {\bfseries 0.639 (0.009)} & {\bfseries 0.568 (0.016)} \\
            GMM SFS F1mean &  {\bfseries 0.767 (0.002)} & {\bfseries 0.646 (0.004)} & {\bfseries 0.768 (0.005)} & {\bfseries 0.605 (0.003)} & {\bfseries 0.634 (0.005)} & {\bfseries 0.569 (0.008)} \\
            GMM SFFS JM &     0.725 (0.025) & 0.528 (0.022) & 0.750 (0.029) & 0.514 (0.023) & 0.585 (0.026) & 0.467 (0.020) \\
            GMM SFFS divKL &  0.408 (0.037) & 0.237 (0.051) & 0.466 (0.020) & 0.241 (0.045) & 0.292 (0.072) & 0.195 (0.048) \\
            GMM SFFS kappa &  0.601 (0.087) & 0.431 (0.065) & 0.641 (0.125) & 0.439 (0.069) & 0.467 (0.081) & 0.373 (0.071) \\
            GMM SFFS OA &     0.544 (0.097) & 0.392 (0.063) & 0.579 (0.127) & 0.396 (0.064) & 0.473 (0.087) & 0.369 (0.066) \\
            GMM SFFS F1mean & 0.473 (0.043) & 0.364 (0.027) & 0.491 (0.061) & 0.361 (0.025) & 0.462 (0.028) & 0.355 (0.023) \\
            \bottomrule
        \end{tabular}
    \end{table}

### 2017-01-13 paper.tex
Mathieu's corrections

\author{Adrien~Lagrange,~Mathieu~Fauvel~and~Manuel~Grizonnet% <-this % stops a space
\thanks{A. Lagrange and M. Fauvel are with the Universit\'{e} de Toulouse,
INP-ENSAT, UMR 1201 DYNAFOR, France and with the INRA, UMR 1201
DYNAFOR, France.}%
\thanks{M. Grizonnet is with Centre National d'\'{E}tudes Spatiales, French Space Agency, Toulouse, France.}%
\thanks{This  work was  supported  by the  French National  Research Agency  (ANR)  under  Project Grant  ANR-13-JS02-0005-01  (Asterix project).}}

@@

The volume of the data is  increasing dramatically with respect to the
number of measurement  per pixel. The data volume  of an hyperspectral
image  is  typically several  hundreds  of  Gigabytes per  acquisition
($\approx$  300km$^2$). Multitemporal  data are  now available  freely
from  internet  stream  (see  for instance  the  Copernicus  data  hub
\url{https://cophub.copernicus.eu/}).  This very  large volume of data
requires   specific   computing  infrastructure.    High   performance
computing   is   actually   investigated   by   the   remote   sensing
community~\cite{christophe2011remote,plaza2011high}.  Main  issues are
related to the use of parallel approaches (multi-core, GPU, clusters) to
improve the  processing time, and  to the use of  streaming techniques
when data  does not fit in  memory.  One popular open  source software
solution  is  the Orfeo  Toolbox,  developed  by the  French  Space
Agency~\cite{christophe2008orfeo}.   Streaming and  parallel computing
are  conveniently proposed  to  users/developpers through  several ``\emph{ready to use}'' modules.

@@

    \begin{figure}[!t]
        \centering
        \begin{tikzpicture}
          \begin{axis}[ymode=log,xmin=0,xmax=200,ymin=0,width=0.8\columnwidth,grid,axis x line=left ,axis y line=left, tick align=outside,xlabel = $d$,ylabel = $\eta_c$,small]
            \addplot+[very thick,mark=none,smooth,domain=0:200,samples=200] (\x,{\x*(\x+3)/2+1});
            \end{axis}
        \end{tikzpicture}
        \caption{Number of parameters $\eta_c$ by class in function of dimension $d$: $\eta_c=d(d+3)/2+1$.\label{fig:nb-param}}
    \end{figure}

@@

        The    KL   divergence    is   not    symmetric,   \emph{i.e.}
        $KL_{cc^\prime} \ne KL_{c^\prime c}$.  A symmetrical version  is used to
        compute the criterion function:
        \begin{align}\label{eq:skl}
          SKL_{cc^\prime} &=KL_{cc^\prime} + KL_{c^\prime c} \nonumber \\
            &= \frac{1}{2} \biggl\{ \text{Tr} (\boldsymbol{\Sigma}_c^{-1} \boldsymbol{\Sigma}_{c^\prime} + \boldsymbol{\Sigma}_{c^\prime}^{-1} \boldsymbol{\Sigma}_c) \nonumber \\
            &~~+ (\boldsymbol{\mu}_c - \boldsymbol{\mu}_{c^\prime})^t (\boldsymbol{\Sigma}_c^{-1} + \boldsymbol{\Sigma}_{c^\prime}^{-1}) (\boldsymbol{\mu}_c - \boldsymbol{\mu}_{c^\prime}) - 2d \biggr\}.
        \end{align}

@@

        The \emph{Jeffries-Matusita distance} is a measure based on the Bhattacharyya distance. It saturates if the separability between the two distributions increases~\cite{bruzzone2009novel}. The JM distance is defined as
        \begin{equation}\label{eq:jm}
            {JM}_{cc^\prime} = \sqrt{ 2 \{1 - \text{exp}(-B_{cc^\prime})\} }.
        \end{equation}

@@

        In the remaining of the paper, $\boldsymbol{\Sigma}_c^{(k-1)}$
        denotes the  covariance matrix of the  $(k-1)^{th}$ iteration,
        \emph{i.e.}, the  covariance matrix  of the  selected features
        and $\boldsymbol{\Sigma}_c^{(k)}$ denotes  a covariance matrix
        at the $k^{th}$ iteration,  \emph{i.e.}, the covariance matrix
        of  an  augmented  set  by  the  feature  $x_k$.  Then,  since
        $\boldsymbol{\Sigma}_c^{(k)}$ is a positive definite symmetric
        matrix, the covariance matrix can be written as

@@

            GMM SFS kappa & 257             & 496             & 955             & {\bfseries 5.2} & {\bfseries 5.2} & {\bfseries 5.5} & 11.95 & 12    & 12.05\\
            GMM SFS JM &    {\bfseries 8.6} & {\bfseries 8.9} & {\bfseries 9.1} & {\bfseries 5.7} & {\bfseries 5.7} & {\bfseries 5.9} & 11.95 & 12    & 12.05\\
            GMM SFFS JM &   {\bfseries 8.8} & {\bfseries 9.0} & {\bfseries 9.3} & {\bfseries 5.0} & {\bfseries 5.0} & {\bfseries 5.4} & 21.45 & 24.35 & 27.05\\
            GMM ridge &     71.7            & 105             & 167             & 530 & 530 & 530    & & & \\
            KNN &           {\bfseries 8.9} & 19.6            & 59.7            & 387 & 639 & 887    & & & \\
            Random Forest & 24.5            & 49.3            & 105             & 33.0 & 41.7 & 45.9 & & & \\
            \bottomrule
        \end{tabularx}
    \end{table*}

>>>>

\author{Adrien~Lagrange ̃\IEEEmembership{Student, IEEE},~Mathieu~Fauvel~\IEEEmembership{Senior, IEEE}~and~Manuel~Grizonnet% <-this % stops a space
\thanks{A. Lagrange and M. Fauvel are with the Universit\'{e} de Toulouse,
INP-ENSAT, UMR 1201 DYNAFOR, France and with the INRA, UMR 1201
DYNAFOR, France.}%
\thanks{M. Grizonnet is with Centre National d'\'{E}tudes Spatiales, French Space Agency, Toulouse, France.}%
\thanks{This  work was  supported  by the  French National  Research Agency  (ANR)  under  Project Grant  ANR-13-JS02-0005-01  (Asterix project).}}

@@

The volume of the data is  increasing dramatically with respect to the
number of measurement  per pixel. The data volume  of an hyperspectral
image  is  typically several  hundreds  of  Gigabytes per  acquisition
($\approx$  300km$^2$). Multitemporal  data are  now available  freely
from  internet  stream  (see  for instance  the  Copernicus  data  hub
\url{https://cophub.copernicus.eu/}).  This very  large volume of data
requires   specific   computing  infrastructure.    High   performance
computing   is   actually   investigated   by   the   remote   sensing
community~\cite{christophe2011remote,plaza2011high}.  Main  issues are
related to the use of parallel approaches (multi-core, GPU, clusters) to
improve the  processing time, and  to the use of  streaming techniques
when data  does not fit in  memory.  One popular open  source software
solution  is  the Orfeo  Toolbox,  developed  by the  French  Space
Agency (CNES)~\cite{christophe2008orfeo}.   Streaming and  parallel computing
are  conveniently proposed  to  users/developpers through  several ``\emph{ready to use}'' modules.

@@

    \begin{figure}[!t]
        \centering
        \begin{tikzpicture}
          \begin{axis}[ymode=log,xmin=0,xmax=200,ymin=0,width=0.8\columnwidth,grid,axis x line=left ,axis y line=left, tick align=outside,xlabel = $d$,ylabel = $\eta_c$,small]
            \addplot+[very thick,mark=none,smooth,domain=0:200,samples=200] (\x,{\x*(\x+3)/2+1});
            \end{axis}
        \end{tikzpicture}
        \caption{Number of parameters $\eta_c$ per class in function of dimension $d$: $\eta_c=d(d+3)/2+1$.\label{fig:nb-param}}
    \end{figure}

@@

        The    KL   divergence    is   not    symmetric,   \emph{i.e.},
        $KL_{cc^\prime} \ne KL_{c^\prime c}$.  A symmetrical version  is used to
        compute the criterion function:
        \begin{align}\label{eq:skl}
          SKL_{cc^\prime} &=KL_{cc^\prime} + KL_{c^\prime c} \nonumber \\
            &= \frac{1}{2} \biggl\{ \text{Tr} (\boldsymbol{\Sigma}_c^{-1} \boldsymbol{\Sigma}_{c^\prime} + \boldsymbol{\Sigma}_{c^\prime}^{-1} \boldsymbol{\Sigma}_c) \nonumber \\
            &~~+ (\boldsymbol{\mu}_c - \boldsymbol{\mu}_{c^\prime})^t (\boldsymbol{\Sigma}_c^{-1} + \boldsymbol{\Sigma}_{c^\prime}^{-1}) (\boldsymbol{\mu}_c - \boldsymbol{\mu}_{c^\prime}) - 2d \biggr\}.
        \end{align}

@@

        The \emph{Jeffries-Matusita distance} is a measure based on the Bhattacharyya distance. It saturates when the separability between the two distributions increases~\cite{bruzzone2009novel}. The JM distance is defined as
        \begin{equation}\label{eq:jm}
            {JM}_{cc^\prime} = \sqrt{ 2 \{1 - \text{exp}(-B_{cc^\prime})\} }.
        \end{equation}

@@

        In the remaining of the paper, $\boldsymbol{\Sigma}_c^{(k-1)}$
        denotes the  covariance matrix of the  $(k-1)^{th}$ iteration,
        \emph{i.e.}, the  covariance matrix  of the  selected features
        and $\boldsymbol{\Sigma}_c^{(k)}$ denotes  a covariance matrix
        at the $k^{th}$ iteration,  \emph{i.e.}, the covariance matrix
        augmented  by  the  feature  $x_k$.  Then,  since
        $\boldsymbol{\Sigma}_c^{(k)}$ is a positive definite symmetric
        matrix, the covariance matrix can be written as

@@

            GMM SFS kappa & 257             & 496             & 955             & {\bfseries 5.2} & {\bfseries 5.2} & {\bfseries 5.5} & 11.95 & 12    & 12.05\\
            GMM SFS JM &    {\bfseries 8.6} & {\bfseries 8.9} & {\bfseries 9.1} & {\bfseries 5.7} & {\bfseries 5.7} & {\bfseries 5.9} & 11.95 & 12    & 12.05\\
            GMM SFFS JM &   {\bfseries 8.8} & {\bfseries 9.0} & {\bfseries 9.3} & {\bfseries 5.0} & {\bfseries 5.0} & {\bfseries 5.4} & 21.45 & 24.35 & 27.05\\
            GMM ridge &     71.7            & 105             & 167             & 530 & 530 & 530    & all & all & all \\
            KNN &           {\bfseries 8.9} & 19.6            & 59.7            & 387 & 639 & 887    & all & all & all \\
            Random Forest & 24.5            & 49.3            & 105             & 33.0 & 41.7 & 45.9 & all & all & all \\
            \bottomrule
        \end{tabularx}
    \end{table*}
